Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression

### 2022-06-27 docs/rmd/presentation/slides.Rmd
autogit

---
title: "Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression"
subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
author: "Roman Hoehn, B.Sc. Wirtschaftspaedagogik"
date: "2022-06-29"
output:
  beamer_presentation:
    colortheme: "beaver"
    fonttheme: "structurebold"
    keep_tex: true
    fig_caption: yes
    number_sections: true
    slide_level: 2
toc: yes
bibliography: ../bib/bib/ref.bib
csl: ../bib/csl/harvard1.csl
header-includes:
   - \usepackage{float}
---

@@

* A metaheuristic that searches for computer programs that solve a given problem
* Inventor: John R. Koza ^[@koza_main]
* Evolutionary algorithm that simulates the process of Darwinian evolution:
  1. Population based
  2. The quality of solutions is evaluated by a fitness function
  3. Selection: Solutions are selected based on their individual fitness
  4. Variation: Mutation and recombination of solutions
* Unique Features:
  * Evolve solutions of variable length and structure
  * Solutions are typically represented by recursive tree structures

@@

* Operator that selects individual solutions from the population for reproduction and mutation
* Most commonly used selection operator in Genetic Programming (GP): Tournament selection^[@10.1007/978-3-642-16493-4_19, p.181]
* Intuition: High chance for "generalist" solutions to be selected since it is based on aggregated fitness scores

## epsilon-Lexicase Selection

* Recent alternative: Lexicase Selection and it's variation $\epsilon$-lexicase selection
* Idea: Selection method for uncompromising, continous-valued symbolic regression problems ^[@6920034, p.12]
* Increases genetic diversity inside the population^[@6920034, p.1]
* Higher chance for "specialist" solutions to be selected since it is decided on a per case basis
* Performance increases have been demonstrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]

## Symbolic Regression

@@

## Generalization

* The ability of a model to perform well on previously unseen fitness cases
* Main objective in most supervised machine learning problems
* Challenge: Avoid overfitting to training data

## Motivation

* Little attention has been paid to generalization in GP ^[@open_issues_gp, @generalisation_in_gp]
* High practical relevance of symbolic regression in many fields, e.g. financial forecasting

@@

>>>>

---
title: "Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression"
subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
author: "Roman Hoehn, B.Sc. Wirtschaftspaedagogik"
date: "2022-06-29"
output:
  beamer_presentation:
    fonttheme: "structurebold"
    keep_tex: true
    fig_caption: yes
    number_sections: true
    slide_level: 2
toc: yes
bibliography: ../bib/bib/ref.bib
csl: ../bib/csl/harvard1.csl
header-includes:
   - \usepackage{float}
---

@@

* A metaheuristic that searches for computer programs that solve a given problem^[@koza_main]
* Evolutionary Algorithm: Simulates the process of Darwinian evolution
* Unique Features:
  * Evolve solutions of variable length and structure
  * Solutions are typically represented by recursive tree structures

@@

### Torunament Selection

* Most commonly used selection operator in Genetic Programming (GP)^[@10.1007/978-3-642-16493-4_19, p.181]
* Intuition: High chance for "generalist" solutions to be selected since it is based on aggregated fitness scores

### $\epsilon$-Lexicase Selection

* Objective: Create Selection method for uncompromising, continous-valued symbolic regression problems ^[@6920034, p.12]
* Performance increases have been demonstrated in many benchmarking problems ^[@epsilon_lexicase_main, p.744-745]
* Intuition: Higher chance for "specialist" solutions to be selected since it is decided on a per case basis

## Related Concepts

### Symbolic Regression

@@

### Generalization

* Main objective in most supervised machine learning problems: Achieve good performance for unseen data
* Challenge: Avoid overfitting to training data
* Little attention has been paid to generalization in GP ^[@open_issues_gp, @generalisation_in_gp]

@@

# Appendix

### 2022-06-27 docs/rmd/presentation/slides.tex
....
autogit

\usecolortheme{beaver}

@@

@@

@@

\begin{frame}{Genetic Programming}
\protect\hypertarget{genetic-programming}{}
\begin{itemize}
\tightlist
\item
  A metaheuristic that searches for computer programs that solve a given
  problem
\item
  Inventor: John R. Koza \footnote<.->{Koza (1992)}
\item
  Evolutionary algorithm that simulates the process of Darwinian
  evolution:

  \begin{enumerate}
  \tightlist
  \item
    Population based
  \item
    The quality of solutions is evaluated by a fitness function
  \item
    Selection: Solutions are selected based on their individual fitness
  \item
    Variation: Mutation and recombination of solutions
  \end{enumerate}
\item
  Unique Features:

@@

\begin{frame}{Parent Selection}
\protect\hypertarget{parent-selection}{}
\begin{itemize}
\tightlist
\item
  Operator that selects individual solutions from the population for
  reproduction and mutation
\item
  Most commonly used selection operator in Genetic Programming (GP):
  Tournament selection\footnote<.->{Fang and Li (2010), p.181}
\item
  Intuition: High chance for ``generalist'' solutions to be selected
  since it is based on aggregated fitness scores
\end{itemize}
\end{frame}

\begin{frame}{epsilon-Lexicase Selection}
\protect\hypertarget{epsilon-lexicase-selection}{}
\begin{itemize}
\tightlist
\item
  Recent alternative: Lexicase Selection and it's variation
  \(\epsilon\)-lexicase selection
\item
  Idea: Selection method for uncompromising, continous-valued symbolic
  regression problems \footnote<.->{Helmuth, Spector and Matheson
    (2015), p.12}
\item
  Increases genetic diversity inside the population\footnote<.->{Helmuth,
    Spector and Matheson (2015), p.1}
\item
  Higher chance for ``specialist'' solutions to be selected since it is
  decided on a per case basis
\item
  Performance increases have been demonstrated in many benchmarking
  problems \footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
\end{itemize}
\end{frame}

\begin{frame}{Symbolic Regression}
\protect\hypertarget{symbolic-regression}{}
\begin{itemize}
\tightlist
\item
  Task: Find a mathematical model that fits a given set of datapoints
\item
  One of the first applications of GP described by Koza (1992)
\item
  High relevance: GP can outperform state-of-the-art machine learning
  algorithms like gradient boosting \footnote<.->{Orzechowski, Cava and
    Moore (2018)}
\end{itemize}
\end{frame}

\begin{frame}{Generalization}
\protect\hypertarget{generalization}{}
\begin{itemize}
\tightlist
\item
  The ability of a model to perform well on previously unseen fitness
  cases
\item
  Main objective in most supervised machine learning problems
\item
  Challenge: Avoid overfitting to training data
\end{itemize}
\end{frame}

\begin{frame}{Motivation}
\protect\hypertarget{motivation}{}
\begin{itemize}
\tightlist
\item
  Little attention has been paid to generalization in GP \footnote<.->{O'Neill
    \emph{et al.} (2010), Kushchu (2002)}
\item
  High practical relevance of symbolic regression in many fields,
  e.g.~financial forecasting
\end{itemize}
\end{frame}

@@

\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
\centering
\begin{tabular}[t]{l|l}
\hline
Variable & Description\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
y1 & Heating Load\\
\hline
y2 & Cooling Load\\
\hline
\end{tabular}
\end{table}
\end{frame}

@@

@@

\caption{\label{tab:unnamed-chunk-2}Evolutionary Parameters}
\centering
\begin{tabular}[t]{l|l}
\hline
Parameter & Value\\
\hline
Population Size & 500\\
\hline
Number of Generations & 100\\
\hline
Mutation Rate & 20\%\\
\hline
Crossover Rate & 80\%\\
\hline
Tournament Size & 3\\
\hline
Epsilon selection & automatic\\
\hline
Elite Size & 0\\
\hline
\end{tabular}
\end{table}
\end{frame}

\begin{frame}{Function Set}
\protect\hypertarget{function-set}{}
\begin{tabular}{l|r}
\hline
Function & Arity\\
\hline
Addition & 2\\
\hline
Subtraction & 2\\
\hline
Multiplication & 2\\
\hline
Negation & 1\\
\hline
Sine & 1\\
\hline
Cosine & 1\\
\hline
Protected Division & 2\\
\hline
Protected Natural Logarithm & 1\\
\hline
Protected Square Root & 1\\
\hline
\end{tabular}
\end{frame}

\begin{frame}{Terminal Set}
\protect\hypertarget{terminal-set}{}
\begin{tabular}{l|l}
\hline
Terminal & Description\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
random\_int & Ephemeral Constant (integer)\\
\hline
random\_float & Ephemeral Constant(float)\\
\hline
\end{tabular}
\end{frame}

@@

\begin{block}{3 Questions}
\protect\hypertarget{questions}{}
\begin{enumerate}
\tightlist
\item
  Differences in average fitness for both algorithms?
\item
  Differences in fitness for training and testing data?
\item
  Differences in program size?
\end{enumerate}
\end{block}
\end{frame}

\begin{frame}{Finding 2}
\protect\hypertarget{finding-2}{}
\begin{itemize}
\tightlist
\item
  The gap between training and testing error is not statistically
  significant for both selection algorithms
\item
  Both algorithms achieve a slightly better performance for the training
  data
\item
  No proof for overfitting
\end{itemize}
\end{frame}

\caption{\label{tab:unnamed-chunk-5}Mean Error - P-Values (MWU)}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
  & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
\midrule
tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
\bottomrule
\end{tabular}
\end{table}
\end{frame}

\begin{frame}{Finding 3}
\protect\hypertarget{finding-3}{}
\ldots{}
\end{frame}

\begin{frame}{Conclusions}
\ldots{}
\end{frame}

\begin{frame}{Limitations and open Questions}
\ldots{} \newpage

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Dua:2019}{}}%
Dua, D. and Graff, C. (2017) {``{UCI} machine learning repository.''}
University of California, Irvine, School of Information; Computer
Sciences. Available at: \url{http://archive.ics.uci.edu/ml}.

@@

>>>>

@@

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
\usepackage{caption}
% Make caption package work with longtable
\def\fnum@table{\tablename~\thetable}

@@

\usepackage{float}

@@

\begin{frame}{Genetic Programming}
\protect\hypertarget{genetic-programming}{}
\begin{itemize}
\tightlist
\item
  A metaheuristic that searches for computer programs that solve a given
  problem\footnote<.->{Koza (1992)}
\item
  Evolutionary Algorithm: Simulates the process of Darwinian evolution
\item
  Unique Features:

@@

\begin{frame}{Parent Selection}
\protect\hypertarget{parent-selection}{}
\begin{block}{Torunament Selection}
\protect\hypertarget{torunament-selection}{}
\begin{itemize}
\tightlist
\item
  Most commonly used selection operator in Genetic Programming
  (GP)\footnote<.->{Fang and Li (2010), p.181}
\item
  Intuition: High chance for ``generalist'' solutions to be selected
  since it is based on aggregated fitness scores
\end{itemize}
\end{block}

\begin{block}{\(\epsilon\)-Lexicase Selection}
\protect\hypertarget{epsilon-lexicase-selection}{}
\begin{itemize}
\tightlist
\item
  Objective: Create Selection method for uncompromising,
  continous-valued symbolic regression problems \footnote<.->{Helmuth,
    Spector and Matheson (2015), p.12}
\item
  Performance increases have been demonstrated in many benchmarking
  problems \footnote<.->{La Cava, Spector and Danai (2016), p.744-745}
\item
  Intuition: Higher chance for ``specialist'' solutions to be selected
  since it is decided on a per case basis
\end{itemize}
\end{block}
\end{frame}

\begin{frame}{Related Concepts}
\protect\hypertarget{related-concepts}{}
\begin{block}{Symbolic Regression}
\protect\hypertarget{symbolic-regression}{}
\begin{itemize}
\tightlist
\item
  Task: Find a mathematical model that fits a given set of datapoints
\item
  One of the first applications of GP described by Koza (1992)
\item
  High relevance: GP can outperform state-of-the-art machine learning
  algorithms like gradient boosting \footnote<.->{Orzechowski, Cava and
    Moore (2018)}
\end{itemize}
\end{block}

\begin{block}{Generalization}
\protect\hypertarget{generalization}{}
\begin{itemize}
\tightlist
\item
  Main objective in most supervised machine learning problems: Achieve
  good performance for unseen data
\item
  Challenge: Avoid overfitting to training data
\item
  Little attention has been paid to generalization in GP \footnote<.->{O'Neill
    \emph{et al.} (2010), Kushchu (2002)}
\end{itemize}
\end{block}
\end{frame}

@@

\begin{longtable}[]{@{}ll@{}}
\caption{Overview - Energy Heating data set}\tabularnewline
\toprule()
Variable & Description \\
\midrule()
\endfirsthead
\toprule()
Variable & Description \\
\midrule()
\endhead
X1 & Relative Compactness \\
X2 & Surface Area \\
X3 & Wall Area \\
X4 & Roof Area \\
X5 & Overall Height \\
X6 & Orientation \\
X7 & Glazing Area \\
X8 & Glazing Area Distribution \\
y1 & Heating Load \\
y2 & Cooling Load \\
\bottomrule()
\end{longtable}
\end{frame}

@@

\begin{block}{Statistical Testing Strategy}
\protect\hypertarget{statistical-testing-strategy}{}
\begin{enumerate}
\tightlist
\item
  Differences in average fitness for both algorithms on both datasets?
\item
  Differences in fitness for training and testing data?
\end{enumerate}
\end{block}
\end{frame}

\begin{frame}{Finding 2}
\protect\hypertarget{finding-2}{}
\begin{itemize}
\tightlist
\item
  The gap between training and testing error is not statistically
  significant for both selection algorithms
\item
  Both algorithms achieve a slightly better performance for the training
  data
\item
  Good generalization: No evidence of overfitting
\end{itemize}
\end{frame}

\caption{\label{tab:unnamed-chunk-2}Mean Error - P-Values (MWU)}
\centering
\fontsize{4}{6}\selectfont
\begin{tabular}[t]{l|r|r|r|r}
\hline
X & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
\hline
tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
\hline
tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
\hline
elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
\hline
elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
\hline
\end{tabular}
\end{table}
\end{frame}

\begin{frame}{Program Growth}
\protect\hypertarget{program-growth}{}
\begin{itemize}
\tightlist
\item
  So far: No proof of differences in generalization
\item
  New approach: Program growth as a possible indicator for overfitting?
\item
  Theory: Minimum description length principle (MDLP) \footnote<.->{Wang,
    Wagner and Rondinelli (2019), p.~268}
\item
  Downside: Growth/Bloat is no clear indicator of overfitting
  \footnote<.->{Silva and Vanneschi (2009), p.~8}
\end{itemize}
\end{frame}

\begin{frame}{Finding 3}
\protect\hypertarget{finding-3}{}
\begin{itemize}
\tightlist
\item
  GP typical growth behaviour for both operators
\item
  Solutions grow at a similiar rate in each generation
\item
  No statistically significant differences in overall program size based
  on selection
\end{itemize}
\end{frame}

\begin{frame}{Conclusions}
\begin{itemize}
\tightlist
\item
  Experiment did not yield evidence for differences in generalization
  behavior between tournament and \(\epsilon\)-lexicase selection
\item
  The performance of tournament selection is significantly higher than
  that of \(\epsilon\)-lexicase selection for the selected symbolic
  regression problem
\item
  No evidence for differences in growth behavior between both algorithms
\end{itemize}
\end{frame}

\begin{frame}{Limitations and open Questions}
\begin{enumerate}
\tightlist
\item
  Configuration of evolutionary parameters
\item
  Results are based on a single symbolic regression
\item
  Limited by computational resources
\end{enumerate}
\end{frame}

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

@@

\caption{\label{tab:unnamed-chunk-3}Evolutionary Parameters}
\centering
\begin{tabular}[t]{l|l}
\hline
Parameter & Value\\
\hline
Population Size & 500\\
\hline
Number of Generations & 100\\
\hline
Mutation Rate & 20\%\\
\hline
Crossover Rate & 80\%\\
\hline
Tournament Size & 3\\
\hline
Epsilon selection & automatic\\
\hline
Elite Size & 0\\
\hline
\end{tabular}
\end{table}
\end{frame}

\begin{frame}[allowframebreaks]{Primitive Set}
\protect\hypertarget{primitive-set}{}
\begin{table}

\caption{\label{tab:unnamed-chunk-4}Function Set}
\centering
\begin{tabular}[t]{l|r}
\hline
Function & Arity\\
\hline
Addition & 2\\
\hline
Subtraction & 2\\
\hline
Multiplication & 2\\
\hline
Negation & 1\\
\hline
Sine & 1\\
\hline
Cosine & 1\\
\hline
Protected Division & 2\\
\hline
Protected Natural Logarithm & 1\\
\hline
Protected Square Root & 1\\
\hline
\end{tabular}
\end{table}
\framebreak

\caption{\label{tab:unnamed-chunk-5}Terminal Set}
\centering
\begin{tabular}[t]{l|l}
\hline
Terminal & Description\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
random\_int & Ephemeral Constant (integer)\\
\hline
random\_float & Ephemeral Constant(float)\\
\hline
\end{tabular}
\end{table}
\end{frame}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\begin{frame}[allowframebreaks]{References}
\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Dua:2019}{}}%
Dua, D. and Graff, C. (2017) {``{UCI} machine learning repository.''}
University of California, Irvine, School of Information; Computer
Sciences. Available at: \url{http://archive.ics.uci.edu/ml}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-bloat_overfitting_gp}{}}%
Silva, S. and Vanneschi, L. (2009) {``Operator equalisation, bloat and
overfitting: A study on human oral bioavailability prediction,''} in
\emph{Proceedings of the 11th Annual Genetic and Evolutionary
Computation Conference, GECCO-2009}, pp. 1115--1122.
doi:\href{https://doi.org/10.1145/1569901.1570051}{10.1145/1569901.1570051}.

\leavevmode\vadjust pre{\hypertarget{ref-wang_wagner_rondinelli_2019}{}}%
Wang, Y., Wagner, N. and Rondinelli, J.M. (2019) {``Symbolic regression
in materials science,''} \emph{MRS Communications}, 9(3), pp. 793--805.
doi:\href{https://doi.org/10.1557/mrc.2019.85}{10.1557/mrc.2019.85}.
